Incorporating periodic variability in hidden Markov models for animal movement

### 2016-07-07 paper3.Rnw
included viterbi inside the plots

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(grepl("HMM",model),"no","yes")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                    , y=deltaBIC
                                    , colour=model
                                    , linetype=model))
  + facet_wrap(~ type,ncol=4)
  + scale_size_continuous( name="# of parameters")
  + scale_x_continuous(expand=c(0,0.5),breaks=2:7)
  + geom_point(aes(size=parameters))
  + scale_linetype_manual(values=linetypev)
  + scale_colour_brewer(palette="Dark2")
  + labs(y=expression(paste(Delta, "BIC"))
         , x="Number of States") 
  + geom_line() 
  + zmargin)
@

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day. \\textbf{todo: leave out hmm quad for clarity? fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such. Handle obs data as in Guelph talk. Make FMM lines dashed use manual linetype scale)})",fig.width=8>>=
# simdf <- simdat %>% dplyr::select(-obs) %>%
#     group_by(time) %>% summarise_each(funs(mean))
# obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
#     filter(!is.na(obs)) %>% group_by(time) %>%
#     summarise_each(funs(mean))
# 
# avgdf2 <- melt(simdf,'time')
avgdf <- readRDS("summary_stats/cat.1.df.RDS")$avgplotdf
obsdf <- (readRDS("summary_stats/cat.1.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)

@@

combdf <- (rbind(avgdf,obsdf)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- rep(c(2,1),c(2,6))
colvec <- c(brewer.pal(7,"Dark2"),"black")
lwdvec <- rep(c(1,4),c(7,1))
alphavec <- rep(c(1,0.2),c(7,1))
(ggplot(combdf,aes(x=time,y=steplen,colour=model,lwd=model,alpha=model,linetype=model)) 
  + geom_line()
  + xlab("Time of Day") 
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_color_manual(values=colvec)
  + scale_alpha_manual(values=alphavec)
  + scale_size_manual(values=lwdvec)
  + scale_linetype_manual(values=linetypev)
)
@

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF \\textbf{fix similarly to Fig 3}">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
# temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)

acfdf <- (readRDS("summary_stats/cat.1.df.RDS")$acfdf 
          %>% mutate(model = .id)
)
(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,lwd=model,alpha=model,linetype=model)) 
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_color_manual(values=colvec)
  + scale_alpha_manual(values=alphavec)
  + scale_size_manual(values=lwdvec)
  + scale_linetype_manual(values=linetypev)
)
@

>>>>

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                    , y=deltaBIC
                                    , colour=model
                                    , linetype=model
                                    , group=model))
  + facet_wrap(~mod,ncol = 2,scales = "free_x")
  + scale_size_continuous(name="# of states")
  + scale_x_continuous(breaks=seq(0,300,by=10))
  + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
  + scale_linetype_manual(values=linetypev)
  + scale_colour_brewer(palette="Dark2")
  + labs(y=expression(paste(Delta, "BIC"))
         , x="Number of States") 
  + geom_line()
  + zmargin)
@

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day. \\textbf{todo: leave out hmm quad for clarity? fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such. Handle obs data as in Guelph talk. Make FMM lines dashed use manual linetype scale)})",fig.width=8>>=
# simdf <- simdat %>% dplyr::select(-obs) %>%
#     group_by(time) %>% summarise_each(funs(mean))
# obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
#     filter(!is.na(obs)) %>% group_by(time) %>%
#     summarise_each(funs(mean))
# 
# avgdf2 <- melt(simdf,'time')
avgdf <- readRDS("summary_stats/cat.1.df.RDS")$avgplotdf
obsdf <- (readRDS("summary_stats/cat.1.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("summary_stats/cat.1.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

combdf <- (rbind(avgdf,obsdf)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6)),1)
colvec <- c(c(brewer.pal(7,"Dark2"),"black"),"black")
lwdvec <- c(rep(c(1,4),c(7,1)),4)
alphavec <- c(rep(c(1,0.2),c(7,1)),1)
(ggplot(combdf,aes(x=time,y=steplen,colour=model,lwd=model,alpha=model,linetype=model)) 
  + geom_line()
  + xlab("Time of Day") 
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_color_manual(values=colvec)
  + scale_alpha_manual(values=alphavec)
  + scale_size_manual(values=lwdvec)
  + scale_linetype_manual(values=linetypev)
) + geom_point(data=vitdf,aes(y=viterbi),lwd=2.5,alpha=1,colour="black",shape=118)
@

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF \\textbf{fix similarly to Fig 3}">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

acfdf <- (readRDS("summary_stats/cat.1.df.RDS")$acfdf 
          %>% mutate(model = .id)
)
(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,lwd=model,alpha=model,linetype=model)) 
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_color_manual(values=colvec)
  + scale_alpha_manual(values=alphavec)
  + scale_size_manual(values=lwdvec)
  + scale_linetype_manual(values=linetypev)
) + geom_point(data=vitacf,lwd=2.5,alpha=1,colour="black",shape=118)
@
